Viral and Bacterial Communities of Colorectal Cancer

### 2017-06-10 doc/manuscript.Rmd
Updated with edits and SRA

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Its risk has been linked to changes in colonic bacterial community composition. Viruses are another important component of the colonic microbial community, however they have yet to be studied in colorectal cancer despite their oncogenic potential. We evaluated the colorectal cancer virome (virus community) in stool using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community. We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer <!--not sure what this is trying to say...guessing no detectable difference in virome a and b diversity between cancer and healthy samples?-->, but more sophisticated random forest models identified striking changes in the virus community. The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function. Our data suggested that the influential phages did not exclusively infect influential bacteria, but rather acted through the community as a whole. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.

@@

# Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g., cancer). Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The strong association of bacterial communities with colorectal cancer and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.

@@

Colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e. adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma may develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr]. Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Thus work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.

@@

Each extraction was performed with a blank buffer control to detect contaminants from reagents or other unintentional sources. Only one of the nine controls contained detectable DNA at a minimal concentration of 0.011 ng/µl, thus providing evidence of the enrichment and purification of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls yielded few sequences and were almost entirely removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**. The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1 kb **(Figure \ref{contigqc})**<!-- scale/notation units in A and C axes don't match; same for B and D x-axes; it would also be awesome if you could label directly on the plot which is viral and which is whole-->. Because contigs only represent genome fragments, we further clustered related bacterial contigs into operational genomic units (OGUs) and viral contigs into operational viral units (OVUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})** to approximate organismal units.

## Unaltered Virome Diversity in Colorectal Cancer
Microbiome and disease associations often are described as being of an altered diversity (i.e. "dysbiotic"). Therefore, first we evaluated the influence of colorectal cancer on virome OVU diversity. We evaluated differences in communities between disease states using the Shannon entropy, richness, and Bray-Curtis metrics. We observed no significant alterations in either Shannon entropy or richness in the diseased state as compared to the healthy state **(Figure \ref{betaogu} C-D)**. There was no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**<!--would be pretty nice/logical to have the treatment colors consistent across panels-->). Notably, there was a significant difference between the few blank controls that remained after rarefying the data and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})** <!--I'd leave out panel A here, it doesn't add anything/is distracting-->, further supporting the quality of the sample set. In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**. This is consistent with what has been observed when the same metrics are applied to 16S rRNA gene sequences (from the same samples) [@Zackular:2014fba] and metagenomic samples [@Zeller:2014ix] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.<!--am I interpretting/re-interpretting this correctly?-->

@@

To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models for classifying stool samples as belonging to individuals with either cancerous or healthy colons. We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]<!--need to more specifically label Fig 1 components in the legend. Maybe as Bacterial 16S, Virome, combined 16S+virome, whole metagenome; the inconsistency is that the "Virus" is metagenome too; maybe label Panel E as "Combined Model IDs"...microbe is a bit ambiguous, not intuitively interpretted as bacteria+virus; could make C and D consistent as "Bacterial Model ID", "Viral Model ID"-->. We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance. The viral model performed as well as the bacterial model (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"])`), with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = asigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = asigfig)`, respectively **(Figure \ref{predmodel} A - B)**.

To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data. The combined model yielded a statistically significant, but minor, performance improvement beyond the viral (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"])`) and bacterial (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"])`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = asigfig)` **(Figure \ref{predmodel} A - B)**. The combined features from the virus and bacterial communities improved our ability to classify stool as belonging to individuals with cancerous colons.

To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed worse than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = asigfig)`) **(Figure \ref{predmodel} A - B)**. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**. Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**. Although the features in the viral model also were of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.

The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was removed. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which was consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature also was driven by a few OVUs, suggesting a role for these viruses in cancer development **(Figure \ref{predmodel} D)**. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and an "unclassified" phage taxon<!--"orphan" is a co-opted term, better to just leave is as unclassified-->. Many of the important viruses were unidentifiable (denoted "unknown"). This is common in viromes across habitats; studies can have as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

@@

After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states. The 16S rRNA gene model model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = asigfig)` and outperformed the viral community model, which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = asigfig)` (p-value `r pvalcalc(threesig)`, **Figure \ref{threewaymodel} A-C**). The microbes important for the cancer versus healthy and healthy versus adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium in the two and three class models was the same *Fusobacterium* (OTU 4) **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. The viruses most important to the three-class model were identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**, but not all important OVUs were of increased abundance in the diseased state **(Figure \ref{threewaymodel} F)**.

## Bacteriophage Dominance in Colorectal Cancer Virome
Changes in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**<!--however, remember the role that the chloroform treatment had on altering the E:B virus ratio...as there are more enveloped eukaryotic viruses than bacterial viruses, these will have lipid bilayers in their capsid and will have been disrupted and lost during the purification step-->. Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes.

We evaluated whether the phages in the community were primarily lytic (obligately lyse their hosts after replication) or temperate (able to integrate into their host's genome to form a lysogen, as well as to transition to a lytic mode). We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky]. We found that the majority of the colon phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**. Thus, the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that <!--we should be careful here not to propogate the false notion that intergrated phages are dormant...quite the contrary, they may be playing a huge role in altering the expression/function of their hosts, potentially more impactful ont he community that lytic phage-->have the ability to integrate into host bacterial genomes. These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages <!--just noting again the potnetial bias that chloroform treatment could play, could simply play down that driven by bacteriopahge versus other viruses was a primariy finding. Could just take this out and start with "We hypotehsized..."-->, we hypothesized that the influential phages were primarily predators of the influential bacteria, and thus influenced their relative abundance through predation. If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacteria and phages. Instead, we observed a strikingly low correlation between bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**. Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential phages were not primarily predators of influential bacteria.

Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. In other words, we hypothesized that the influential bacteriophages were community hubs (central members) within the bacteria and phage interactive network. We investigated the <!--not sure what "infectious capabilities" means here or how it is different from host range/who infects whom--> potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community (**Cite Network Preprint**). The predicted interactions were then used to identify phage community hubs. This revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)** <!-- do you mean high prevalence of phages wide host ranges? or range of possible tropisms (as it technically reads now, I think)...which would mean some pahge infect one host, some infect a hgue number of hosts...if so, what is the max number of hosts predicted, or the distribution of host ranges? This is an interesting feature to report. I'd also rephrase or state expicitly what wide tropism range means with repsect to the network representation.--> . We calculated the alpha centrality (measure of importance in the ecological network) of each phage OVU's connection to the rest of the network. The phages with high centrality values were defined as community hubs. Next, the centrality of each OVU was compared to its importance in the colorectal cancer classification model. Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.

# Discussion
Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial communities [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb]. Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, the link between virus colorectal communities and colorectal cancer has yet to be evaluated. Here we show that, like colonic bacterial communities, the colon virome was altered in colorectal cancer. Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition <!--can you isolate this scenario from the oncogenesis by bacteria-modulated phage commumity composition?-->.

Here, we have begun to delineate the role the colonic virome played in colorectal cancer **(Figure \ref{modelsummary} A)**. We found basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states. By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages <!--again, have to nuance this, biases by chloroform treatment...lipid layer-containing euk viruses are disproportionately more influenced by chlor than phages-->. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host.

>>>>

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Its risk has been linked to changes in colonic bacterial community composition. Viruses are another important component of the colonic microbial community, however they have yet to be studied in colorectal cancer despite their oncogenic potential. We evaluated the colorectal cancer virome (virus community) in stool using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community. We identified no detectable difference in diversity (alpha or beta) between healthy, adenomatous, or cancerous colonic samples, but more sophisticated random forest models identified striking changes in the virus community. The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting that the community was indirectly linked to colorectal cancer by modulating bacterial community structure and function. Our data suggested that the influential phages did not exclusively infect influential bacteria, but rather acted through the community as a whole. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities.

@@

# Introduction
Due to their mutagenic abilities and propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk] and have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb; @Arthur:2012kl], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut, including eukaryotic and prokaryotic viruses) therefore has the potential to impact health and disease (e.g. cancer). Altered human virome composition and diversity have been identified in diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], cystic fibrosis [@Willner:2009dq], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The strong association of bacterial communities with colorectal cancer and the precedent for the virome to impact other human diseases suggest that colorectal cancer may be associated with altered virus communities.

@@

Colorectal cancer progresses in a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e. adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma may develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr]. Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Thus work that aims to facilitate early detection and prevention of progression beyond early cancer stages has great potential to inform therapeutic development.

@@

Each extraction was performed with a blank buffer control to detect contaminants from reagents or other unintentional sources. Only one of the nine controls contained detectable DNA at a minimal concentration of 0.011 ng/µl, thus providing evidence of the enrichment and purification of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls yielded few sequences and were almost entirely removed while rarefying the datasets to a common number of sequences **(Figure \ref{qualcontrol} B)**. The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1 kb **(Figure \ref{contigqc})**. Because contigs only represent genome fragments, we further clustered related bacterial contigs into operational genomic units (OGUs) and viral contigs into operational viral units (OVUs) **(Figure \ref{contigqc} - \ref{clustercontigqc})** to approximate organismal units.

## Unaltered Virome Diversity in Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). We therefore first evaluated the influence of colorectal cancer on virome OVU diversity. We evaluated differences in communities between disease states using the Shannon entropy, richness, and Bray-Curtis metrics. We observed no significant alterations in either Shannon entropy or richness in the diseased states as compared to the healthy state **(Figure \ref{betaogu} C-D)**. There was no statistically significant clustering of the disease groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"])`, **Figure \ref{betaogu}**). Notably, there was a significant difference between the few blank controls that remained after rarefying the data and the other study groups (ANOSIM p-value `r pvalcalc(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"])`, **Figure \ref{betaogunegative})**, further supporting the quality of the sample set. In summary, standard alpha and beta diversity metrics were insufficient for capturing virus community differences between disease states **(Figure \ref{betaogu})**. This is consistent with what has been observed when the same metrics were applied to metagenomic samples [@Zeller:2014ix] and points to the need for alternate approaches to detect the impact of colorectal cancer disease state on these communities.

@@

To identify the altered virus communities associated with colorectal cancer, we built and tested random forest models for classifying stool samples as belonging to individuals with either cancerous or healthy colons. We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]. We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance. The viral model performed as well as the bacterial model (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"])`), with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = asigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = asigfig)`, respectively **(Figure \ref{predmodel} A - B)**.

To evaluate the ability of both bacterial and viral biomarkers to classify samples, we built a combined model that used both bacterial and viral community data. The combined model yielded a modest but statistically significant performance improvement beyond the viral (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Virus"),"pval"])`) and bacterial (corrected p-value `r pvalcalc(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"])`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = asigfig)` **(Figure \ref{predmodel} A - B)**. The combined features from the virus and bacterial communities improved our ability to classify stool as belonging to individuals with cancerous colons.

To determine the advantage of viral metagenomic methods over bacterial metagenomic methods, we compared the viral model to a model built using relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed worse than the other models (mean AUC = `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = asigfig)`) **(Figure \ref{predmodel} A - B)**. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and low abundance OTUs **(Figure \ref{16scompare})**. Removal of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome-based model **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% across the samples **(Figure \ref{16scompare} B)**. Although the features in the viral model also were of low abundance **(Figure \ref{threewaymodel} F)**, the coverage was sufficient for high model performance, likely because viral genomes are orders of magnitude smaller than bacterial genomes.

The association between the bacterial and viral communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was removed. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which was consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature also was driven by a few OVUs, suggesting a role for these viruses in cancer development **(Figure \ref{predmodel} D)**. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and "unclassified" phage taxa. Many of the important viruses were unidentifiable (denoted "unknown"). This is common in viromes across habitats; studies have reported as much as 95% of virus sequences belonging to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz; @Willner:2009dq; @Brum:2015iaa]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

@@

After evaluating our ability to classify samples between two disease states, we performed a three-class random forest model including all disease states. The 16S rRNA gene model yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = asigfig)` and outperformed the viral community model, which yielded a mean AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = asigfig)` (p-value `r pvalcalc(threesig)`, **Figure \ref{threewaymodel} A-C**). The microbes important for the cancer versus healthy and healthy versus adenoma models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium in the two and three class models was the same *Fusobacterium* (OTU 4) **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. The viruses most important to the three-class model were identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**, but not all important OVUs were of increased abundance in the diseased state **(Figure \ref{threewaymodel} F)**.

## Bacteriophage Dominance in Colorectal Cancer Virome
Changes in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that were important for colorectal cancer, we identified the virome OVUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall, we were able to identify `r signif(idcount, digits = sigfig)`% of the OVUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OVUs aligned to bacteriophage reference genomes. It is important to note that this could have been influenced by our methodological biases against enveloped viruses, due to chloroform and DNase treatment for purification.

We evaluated whether the phages in the community were primarily lytic (obligately lyse their hosts after replication) or temperate (able to integrate into their host's genome to form a lysogen, as well as to transition to a lytic mode). We accomplished this by identifying three markers for temperate phages in the OVU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes, as previously described [@Minot:2011ez; @Hannigan:2015fz; @Hannigan:2017ky]. We found that the majority of the colon phages were temperate and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**. Thus, the majority of the OVUs were temperate bacteriophages and not eukaryotic viruses, indicating that the association between the virome and colorectal cancer was reliant on bacteriophage communities that have the ability to integrate into host bacterial genomes. These findings were consistent with previous reports suggesting the gut virome is primarily composed of temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were primarily predators of the influential bacteria, and thus influenced their relative abundance through predation. If this hypothesis were true, we would expect a correlation between the relative abundances of influential bacteria and phages. Instead, we observed a strikingly low correlation between bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**. Overall, there was an absence of correlation between the most influential OVUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential phages were not primarily predators of influential bacteria.

Given these findings, we hypothesized that the most influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the influential bacteria. In other words, we hypothesized that the influential bacteriophages were community hubs (central members) within the bacteria and phage interactive network. We investigated the potential host ranges of all phage OVUs using a previously developed random forest model that relies on sequence features to predict which phages infected which bacteria in the community **(Figure \ref{network} A)** [@Hannigan:2017cj]. The predicted interactions were then used to identify phage community hubs. We calculated the alpha centrality (measure of importance in the ecological network) of each phage OVU's connection to the rest of the network. The phages with high centrality values were defined as community hubs. Next, the centrality of each OVU was compared to its importance in the colorectal cancer classification model. Phage OVU centrality was significantly and positively correlated with importance to the disease model (p-value `r pvalcalc(corstats[c(corstats$names %in% "pval"),"scores"])`, R = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), suggesting that phages important in driving colorectal cancer were also more likely to be community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.

# Discussion
Because of their propensity for mutagenesis and capacity for modulating their host functionality, many viruses are oncogenic [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Some bacteria also have oncogenic properties, suggesting that bacteriophages may play an indirect role in promoting carcinogenesis by influencing bacterial community composition and dynamics [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb]. Despite their carcinogenic potential and the strong association between bacteria and colorectal cancer, the link between virus colorectal communities and colorectal cancer has yet to be evaluated. Here we show that, like colonic bacterial communities, the colon virome was altered in colorectal cancer. Our findings support a working hypothesis for oncogenesis by phage-modulated bacterial community composition.

Here, we have begun to delineate the role the colonic virome played in colorectal cancer **(Figure \ref{modelsummary} A)**. We found basic diversity metrics of alpha diversity (richness and Shannon entropy) and beta diversity (Bray-Curtis dissimilarity) were insufficient for identifying virome community changes between healthy and cancerous states. By implementing a more sophisticated machine learning approach (random forest classification), we detected strong associations between the colon virus community composition and colorectal cancer. The colorectal cancer virome was composed primarily of bacteriophages. These phage communities were not exclusively predators of the most influential bacteria, as demonstrated by the lack of correlation between the abundances of the bacterial and phage populations. Instead, we identified influential phages as being community hubs, suggesting phages influence cancer by altering the greater bacterial community instead of directly modulating the influential bacteria. Our previous work has shown that modifying colon bacterial communities alters colorectal cancer progression and tumor burden in mice [@Zackular:2016en; @Baxter:2014hb]. This provides a precedent for phage indirectly influencing colorectal cancer progression by altering the bacterial community composition. Overall, our data support a model in which the bacteriophage community modulates the bacterial community, and through those interactions indirectly influences the bacteria driving colorectal cancer progression **(Figure \ref{modelsummary} A)**. Although our evidence suggested phages indirectly influenced colorectal cancer development, we were not able to rule out the role of phages directly interacting with the human host.

In addition to modeling the basic potential connections between virus communities, bacteria communities, and colorectal cancer, we also used our data and existing knowledge of phage biology to develop a working hypothesis for the mechanisms by which this may occur. This was done by incorporating our findings into the current model for colorectal cancer development **(Figure \ref{modelsummary} B)** [@Flynn:2016iu]. We hypothesize that the process began with broadly infectious phages in the colon lysing and thereby disrupted the existing bacterial communities. This shift led to novel niche space that enabled opportunistic bacteria (such as *Fusobacterium nucleatum*) to colonize. Once the initial influential founder bacteria established themselves in the epithelium, secondary opportunistic bacteria were able to adhere to the founders, colonize, and begin establishing a biofilm. Phages may have played a role in biofilm dispersal and growth by lysing bacteria within the biofilm, a process important for effective biofilm growth [@Rossmann:2015cj]. The oncogenic bacteria may then have been able to transform the epithelial cells and disrupt tight junctions to infiltrate the epithelium, thereby initiating an inflammatory immune response. As the adenomatous polyps developed and progressed towards carcinogenesis, we observed a shift in the phages and bacteria whose relative abundances were most influential. As the bacteria entered their oncogenic synergy with the epithelium, we conjecture that the phages continued mediating biofilm dispersal. This process would thereby support the colonized oncogenic bacteria by lysing competing cells and releasing nutrients to other bacteria in the form of cellular lysates. In addition to highlighting the likely mechanisms by which the colorectal cancer virome is interacting with the bacterial communities, this outline will guide future research investigations of the role the virome plays colorectal cancer.

A notable finding was the poor performance observed using bacterial metagenomic methods compared to the performance of models using viral metagenomes or 16S rRNA gene sequences. This observation highlights the importance of high sequencing coverage in bacterial metagenomic studies and the advantage of 16S rRNA gene sequencing over whole metagenomic shotgun sequencing <!--do you ever lay out the logic that this is what you think explains this (I don't remember that discussion yet)? if not, should do that first here, briefly-->. Six bacterial OTUs  drove the performance of the 16S rRNA gene classification model and these OTUs were all sparsely present and lowly abundant. Filtration of OTUs with a median relative abundance of zero resulted in the removal of the six important OTUs and reduced model performance to being nearly random like the bacterial metagenomic model. The bacterial metagenomic OGUs represented only the most abundant taxa, which was not informative for this application. There has been some success in using shotgun metagenomic approaches for stool colorectal cancer classification [@Zeller:2014ix], but this previous approach relied on lowly abundant signatures and did not utilize OGU clustering, as done here. In that former case, the models only performed as well as the 16S rRNA gene model [@Zeller:2014ix]. Thus, the targeted 16S rRNA gene sequencing approach, which yielded only a fraction of the bacterial metagenomic sequences, was more effective for detecting colorectal cancer in stool samples. Despite a loss of enthusiasm for 16S rRNA gene sequencing in favor of shotgun metagenomic techniques, 16S rRNA gene sequencing is still a superior methodological approach for some important applications.
